
```{r,warning=FALSE,message=FALSE,error=FALSE, results='hide' , include=FALSE}
knitr::opts_chunk$set( out.width='100%', fig.asp=NULL, fig.align='center', echo=F, warning=FALSE,message=FALSE,error=FALSE, results='hide', cache=T  ) 

#Notes to self
#It's not like R notebook you have the knit every time you want to update html, you can't just save
#You don't have to rerun the code or rebuild the whole book though.

```

```{r}
#Library Loads
library(scales) 
library(tidyverse)
# stable version on CRAN
#install.packages("bookdown")
# or development version on GitHub
# devtools::install_github('rstudio/bookdown')
#libraries
library(lubridate)
library(tidyverse)

#devtools::install_github("ropensci/USAboundaries")
#devtools::install_github("ropensci/USAboundariesData")
library(USAboundaries) ; #install.packages('USAboundaries')
data(state_codes)

library(tidyverse)
library(scales)
library(gghighlight)
library(lubridate)
library(R0)  # consider moving all library commands to top -- this one was in a loop below

library(WikidataR)
library(countrycode)

library(usmap) ; # install.packages('usmap')
data(statepop)
#devtools::install_github("ropensci/USAboundaries")
#devtools::install_github("ropensci/USAboundariesData")
library(USAboundaries) ; #install.packages('USAboundaries')
data(state_codes)

library(tidyverse)
library(sf)

library(jsonlite)

#This is too slow it's downloading each
library(GADMTools)
library(strucchange) ; #install.packages('strucchange')
library(tsibble)

library(patchwork)
```


```{r}
#Data load
lhs_long_clean_labeled <- readRDS( "/media/skynet2/905884f0-7546-4273-9061-12a790830beb/rwd_github_private/NESScovid19/data_temp/lhs_long_clean_labeled.Rds")
dim(lhs_long_clean_labeled) #716545 #704314 #675,708 #187,305

```


# COVID-19 Inference

In this chapter I will introduce parameters we wish to infer about COVID-19 spread and a modeling approach that can recover them from the observable introduced in the last chapter. I will add components to the model one piece at a time so that their function and impact on inference are clear. 

## Confirmed Cases

```{r}

gid_cfr <- lhs_long_clean_labeled %>% 
           arrange(-confirmed) %>% 
           group_by(gid, admin_level, name_prefered) %>% #this doesn't work because there are some with more deaths than confirmed
            filter(row_number()==1) %>%
            #summarize(confirmed=max(confirmed,na.rm=T), deaths=max(deaths,na.rm=T)) %>% 
           ungroup() %>% 
           dplyr::select(gid, admin_level, name_prefered, confirmed, deaths) %>%
           mutate_if(is.numeric, list(~na_if(abs(.), Inf))) %>% distinct() %>%
           na.omit() %>% 
           arrange(-confirmed) %>%
           mutate(cfr=deaths/confirmed) %>%
           filter(cfr<.25) #once you remove these 17 outliers evrything is flagged as beta distributed

library(fitdistrplus)
gid_cfr %>% filter(admin_level==0) %>% pull(confirmed) %>% quantile(prob=seq(0,1,.01))
gid_cfr %>% filter(admin_level==0) %>% pull(confirmed) %>% quantile(prob=seq(0,1,.1))
gid_cfr %>% filter(admin_level==0) %>% pull(confirmed) %>% hist()
gid_cfr %>% filter(admin_level==0) %>% pull(confirmed) %>% fitdistrplus::descdist(boot = 1000)

gid_cfr %>% filter(admin_level==1) %>% pull(confirmed) %>% fitdistrplus::descdist(boot = 1000)

gid_cfr %>% filter(admin_level==1) %>% pull(confirmed) %>% quantile(prob=seq(0,1,.1))
gid_cfr %>% filter(admin_level==2) %>% pull(confirmed) %>% fitdistrplus::descdist(boot = 1000)
gid_cfr %>% filter(admin_level==2) %>% pull(confirmed) %>% hist()
# CHN.13.1_1            2 China, Hubei, Enshi Tujia and Miao     68134   4512 0.0662 #there are a lot of dupes here


```


## Confirmed Case Fatality Rate (CCFR)

One parameter we can estimate directly from the data is the ratio of reported COVID-19 fatalities to the reported COVID-19 cases in an area. This should not be confused with the Case Fatality Rate (CFR) or the Infection Fatality Rate (IFR) which I discuss later. The CCFR is bounded between 0 and 1.

$Confirmed Case Fataility Rate = \frac{ConfirmedCovid19Fatailities}{ConfirmedCovid19Cases}$

```{r}

gid_cfr <- lhs_long_clean_labeled %>% 
           arrange(-confirmed) %>% 
           group_by(gid, admin_level, name_prefered) %>% #this doesn't work because there are some with more deaths than confirmed
            filter(row_number()==1) %>%
            #summarize(confirmed=max(confirmed,na.rm=T), deaths=max(deaths,na.rm=T)) %>% 
           ungroup() %>% 
           dplyr::select(gid, admin_level, name_prefered, confirmed, deaths) %>%
           mutate_if(is.numeric, list(~na_if(abs(.), Inf))) %>% distinct() %>%
           na.omit() %>% 
           arrange(-confirmed) %>%
           mutate(cfr=deaths/confirmed) %>%
           filter(cfr<.25) #once you remove these 17 outliers evrything is flagged as beta distributed

#gid_cfr %>% filter(cfr>.25) #only 17 examples with cfr greater that .25 and it's all low level stuff

summary(gid_cfr$cfr) #99% are less than .25 going to exclude anything over .25
gid_cfr %>% filter(admin_level==0) %>% pull(cfr) %>% quantile(prob=seq(0,1,.01))
gid_cfr %>% filter(admin_level==1) %>% pull(cfr) %>% quantile(prob=seq(0,1,.01))
gid_cfr %>% filter(admin_level==2) %>% pull(cfr) %>% quantile(prob=seq(0,1,.01))

library(fitdistrplus)
gid_cfr %>% filter(admin_level==0) %>% pull(cfr) %>% fitdistrplus::descdist(boot = 1000) #beta distributed
gid_cfr %>% filter(admin_level==1) %>% pull(cfr) %>% fitdistrplus::descdist(boot = 1000) #beta distributed
gid_cfr %>% filter(admin_level==2) %>% pull(cfr) %>% fitdistrplus::descdist(boot = 1000) #beta distributed

fg <- gid_cfr %>% filter(admin_level==0) %>% pull(cfr) %>% fitdist("beta")
par(mfrow = c(2, 2))
plot.legend <- c("Beta")
denscomp(list(fg), legendtext = plot.legend)
qqcomp(list(fg), legendtext = plot.legend)
cdfcomp(list(fg), legendtext = plot.legend)
ppcomp(list(fg), legendtext = plot.legend)

fg <- gid_cfr %>% filter(admin_level==1) %>% pull(cfr) %>% fitdist("beta")
par(mfrow = c(2, 2))
plot.legend <- c("Beta")
denscomp(list(fg), legendtext = plot.legend)
qqcomp(list(fg), legendtext = plot.legend)
cdfcomp(list(fg), legendtext = plot.legend)
ppcomp(list(fg), legendtext = plot.legend)

fg <- gid_cfr %>% filter(admin_level==2) %>% pull(cfr) %>% fitdist("beta")
par(mfrow = c(2, 2))
plot.legend <- c("Beta")
denscomp(list(fg), legendtext = plot.legend)
qqcomp(list(fg), legendtext = plot.legend)
cdfcomp(list(fg), legendtext = plot.legend)
ppcomp(list(fg), legendtext = plot.legend)


global_cfr <- gid_cfr %>% filter(admin_level==0) %>% summarise_if(is.numeric, sum, na.rm=T) %>% mutate(cfr=round(deaths/confirmed,2))

```

$CFR=`r global_cfr$cfr %>% round(2)` = \frac{`r global_cfr$global_cfr %>% comma()`}{`r global_cfr$global_cfr %>% comma()`}$


```{r, eval=F}
gid_cfr %>% mutate(admin_level=as.factor(admin_level)) %>% ggplot(aes(x=cfr, color=admin_level)) + geom_density() + theme_bw()

gid_cfr %>% pull(cfr) %>% summary()

gid_cfr %>% mutate(admin_level=as.factor(admin_level)) %>% ggplot(aes(x=cfr, color=admin_level)) + geom_boxplot() + theme_bw()


```

```{r, fig.height=3, fig.width=8}
library(ggridges)
temp <- gid_cfr %>% 
  mutate(admin_level=as.factor(admin_level)) %>%
  mutate(admin_level=recode(admin_level, `0` = "Country", `1`="Province", `2`="District")) %>%
  mutate(cfr=cfr*100) 
  
temp %>%
  ggplot(aes(x = cfr, y = admin_level)) + #
  geom_density_ridges(scale = 1.6, quantile_lines = TRUE, fill="lightgrey", alpha=.6, aes( vline_color = ..quantile.., vline_type="dotted")
                      #point_shape="|", point_size=.01, size=0.25, jittered_points=T
                      ) + 
  scale_y_discrete(expand = c(0.1, 0)) +     # will generally have to set the `expand` option
  scale_x_continuous(expand = c(0, 0), labels = function(x) paste0(x, "%")) +   # for both axes to remove unneeded padding #https://stackoverflow.com/questions/27433798/how-can-i-change-the-y-axis-figures-into-percentages-in-a-barplot
  coord_cartesian(clip = "off") + # to avoid clipping of the very top of the top ridgeline
  theme_ridges() + coord_cartesian(xlim=c(0,25)) +
  scale_discrete_manual("vline_color",
                      values = c("blue", "red", "blue", "black"), 
                      breaks = c(1, 2),
                      labels = c("1st & 3rd quartile", "median"),
                      name = NULL) +
  theme(legend.position = c(0.6, 0.8)) +
  ylab("Aggregation") + xlab("Confirmed Case Fatality Rate") +
  labs(
    title="Distribution of Confirmed Case Fataility Rate (CCFR)",
    subtitle="Excluding Outliers"
  ) +
  geom_text(data=temp %>% group_by(admin_level) %>% 
            summarise(cfr= round(median(cfr, na.rm=T), 2)),
          aes(label=paste0(round(cfr,1),"%")), 
          position=position_nudge(y=.8, x=.7), colour="red", size=3.5)  +
  geom_text(data=temp %>% group_by(admin_level) %>% 
            summarise(cfr= round(quantile(cfr,probs=.25, na.rm=T), 2)),
          aes(label=paste0(round(cfr,1),"%")), 
          position=position_nudge(y=.8, x=.7), colour="blue", size=3.5) +
  geom_text(data=temp %>% group_by(admin_level) %>% 
            summarise(cfr= round(quantile(cfr,probs=.75, na.rm=T), 2)),
          aes(label=paste0(round(cfr,1),"%")), 
          position=position_nudge(y=.8, x=.7), colour="blue", size=3.5) 

```





## Tests

## Tested People versus Tested Samples

## Interpolate Within Observed

## Interplate Prior to Observed

## Interpolate Subnationally

## Explaining Variation in Testing



South Korea

Vietnam
https://www.reuters.com/article/us-health-coronavirus-vietnam-fight-insi-idUSKBN22B34H



